---
title: "Epi 751 Lab6"
output:
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
---

```{r, echo=FALSE,results=FALSE, warning=FALSE, message=FALSE}
# install.packages("haven", repos="https://cran.r-project.org/web/packages/haven/index.html")
# install.packages("survival", repos="https://cran.r-project.org/web/packages/survival/index.html")
# install.packages("tidyverse", repos="https://cran.r-project.org/web/packages/tidyverse/index.html")
# install.packages("dplyr", repos="https://cran.r-project.org/web/packages/dplyr/index.html")
# install.packages("devtools", repos="https://www.r-project.org/nosvn/pandoc/devtools.html")
# tinytex::install_tinytex(force=TRUE)
library(tidyverse)
library(devtools)
library(dplyr)
library(haven)
library(survival)
library(devtools)
# install_github('kassambara/survminer')
library(survminer)

```

### Read in the data.

```{r, echo=TRUE}
#save the files "Lab6_ARIC_Data.dta" and "Lab6.Rmd" to your desktop or a known folder.
#if files are on your desktop, replace *** with your username for Mac Users. for PC users, insert the filepath to your Desktop or wherever you are storing the Lab6_ARIC_Data file
setwd("D:/JHU/term 1/751-epi method/week 6/Epi751_Lab6_DataFiles") #this line sets the working directory for where files are read from and saved to.

data <- read_dta("Lab6_ARIC_Data.dta")
data$fupchdn<-ifelse(data$fupchd==0,'no_CHD_event','CHD_event') #make a variable that describes CHD events over the followup period using words rather than numbers
```

# Calculate how many participants are included in this ARIC dataset

```{r, echo=TRUE}
length(data$id)
```

# Examine the end dates

```{r, echo=TRUE}
table_enddate <- table(data$enddate)
table_enddate
```

# Explore & Describe the Dataset

## Examine the breakdown of gender data.

```{r, echo=TRUE}
data$gender<-ifelse(data$gender==1,'Male','Female')
table(data$gender, data$fupchdn)
sumgender <- sum(is.na(data$gender))
print(paste0("Number of missing values: ",sumgender))
```

## Examine the breakdown of race data.

```{r, echo=TRUE}
data$racegrp <- as.numeric(data$racegrp)
data$racegrp[data$racegrp==1] <- "White"
data$racegrp[data$racegrp==2] <- "Black"
data$racegrp[data$racegrp==3] <- "Other"
table(data$racegrp, data$fupchdn)
sumrace <- sum(is.na(data$racegrp))
print(paste0("Number of missing values: ",sumrace))
```

## Examine the prevalence of smoking in this cohort.

```{r, echo=TRUE}
data$smoke123 <- as.numeric(data$smoke123)
data$smoke123[data$smoke123==1] <- "Current"
data$smoke123[data$smoke123==2] <- "Former"
data$smoke123[data$smoke123==3] <- "Never"
table(data$smoke123, data$fupchdn)
sumsmoke <- sum(is.na(data$smoke123))
print(paste0("Number of missing values: ",sumsmoke))
```

## Examine hypertension.

```{r, echo=TRUE}
data$hypert <- as.numeric(data$hypert)
data$hypert<-ifelse(data$hypert==1,'Hypertension','No Hypertension')
table(data$hypert, data$fupchdn)
sumrace <- sum(is.na(data$hypert))
print(paste0("Number of missing values: ",sumrace))
```

## Examine cholesterol.

```{r, echo=TRUE}
data$cholcat <- as.numeric(data$cholcat)
data$cholcat[data$cholcat==1] <- "Chol < 200"
data$cholcat[data$cholcat==2] <- "200 < Chol < 240"
data$cholcat[data$cholcat==3] <- "Chol > 240"
table(data$cholcat, data$fupchdn)
sumcholcat <- sum(is.na(data$cholcat))
print(paste0("Number of missing values: ",sumcholcat))
```

## Examine BMI.

```{r, echo=TRUE}
data$bmicat <- as.numeric(data$bmicat)
data$bmicat[data$bmicat==1] <- "Normal"
data$bmicat[data$bmicat==2] <- "Overweight"
data$bmicat[data$bmicat==3] <- "Obese"
table(data$bmicat, data$fupchdn)
sumbmicat <- sum(is.na(data$bmicat))
print(paste0("Number of missing values: ",sumbmicat))
```

\newpage

# Prevalence of Prior CHD

```{r, echo=TRUE}
data$prevchd <- as.numeric(data$prevchd)
data$prevchd<-ifelse(data$prevchd==1,'Yes','No')
table(data$prevchd, data$fupchdn)
sumprevchd <- sum(is.na(data$prevchd))
print(paste0("Number of missing values: ",sumprevchd))
```

# Total follow-up time excluding those with prevalent CHD at baseline

```{r, echo=TRUE}
noprevchd <- data %>%
    filter(prevchd=="No") %>%
    select(id,followup, gender, racegrp, bmicat,smoke123, fupchd)
sum(noprevchd$followup)
```

# Number of failures in single failure-per-subject data

```{r, echo=TRUE}
table(data$fupchdn, data$prevchd)
```

\newpage

# Incidence of CHD by participant characteristics

## By Gender:

```{r gender, echo=TRUE}
aggregate(followup~gender, data=noprevchd,sum) #person-time at risk by gender
aggregate(fupchd~gender, data=noprevchd, sum) #number of incident CHD cases by gender
table(noprevchd$gender) #total number of participants by gender
```

## By Race:

```{r race, echo=TRUE}
aggregate(followup~racegrp, data=noprevchd,sum) #person-time at risk
aggregate(fupchd~racegrp, data=noprevchd,sum) #number of incident CHD cases
table(noprevchd$racegrp) #total number of participants
```

## By BMI:

```{r bmi, echo=TRUE}
aggregate(followup~bmicat, data=noprevchd,sum) #person-time at risk
aggregate(fupchd~bmicat, data=noprevchd,sum) #number of incident CHD cases
table(noprevchd$bmicat) #total number of participants
```

## By Smoking:

```{r}
aggregate(followup~smoke123, data=noprevchd,sum) #person-time at risk
aggregate(fupchd~smoke123, data=noprevchd,sum) #number of incident CHD cases
table(noprevchd$smoke123) #total number of participants
```

\newpage

# Kaplan-Meier Curves

```{r, echo=TRUE}
#these two lines of code preps the survival object needed to make Kaplan-Meier curves
data$fupchd <- as.numeric(data$fupchd) 
survival_object <- Surv(time=noprevchd$followup, event=noprevchd$fupchd)
```

## Making the graphs. For the first two graphs, only focus on the solid line.

```{r, echo=TRUE}
km_fit <- survfit(survival_object ~ 1)
km_gender <- survfit(survival_object ~ gender, data=noprevchd)
plot(km_fit, xlab = "Time in Years", ylab = "Probability of Survival", main="Kaplan-Meier Survival Estimates")
```

```{r, echo=TRUE}
plot(km_fit,ylim=c(0.9,1),xlab = "Time in Years", ylab = "Probability of Survival", main="Kaplan-Meier Survival Estimates")
```

```{r, echo=TRUE}
plot(km_gender, ylim=c(0.9,1),xlab = "Time in Years", ylab = "Probability of Survival", main="Kaplan-Meier Survival Estimates", col=c("blue","red"))
legend(x="topright", legend =c("Female","Male"), fill =c("blue","red"))
```
